Detecting the impact of land cover change on observed rainfall.

### 2019-06-17 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.tex
changed files after review and resubmitted PeerJ

\usepackage{lipsum} \usepackage{textcomp, rotating}

@@

\begin{abstract}
Analysis of observational data to pinpoint impact of land cover change
on local rainfall are difficult due to multiple environmental factors
that cannot be strictly controlled. In this study we use a statistical
approach to identify the relationship between removal of tree cover and
rainfall with data from best available sources for two large areas in
Australia. Gridded rainfall data between 1979 and 2015 was used for the
areas, while large scale (exogenous) effects were represented by mean
rainfall across a much larger area and climatic indicators, such as
Southern Oscillation Index and Indian Ocean Dipole. Both general
additive modelling and step trend tests were used for the analysis. For
a region in south central Queensland, the reported change in tree
clearing between 2002 - 2005 did not result in strong statistically
significant precipitation changes. On the other hand, results from a
bushfire affected region on the border of New South Wales and Victoria
suggests significant changes in the rainfall due to changes in tree
cover. This indicates the method works better when a abrupt change in
the data can be clearly identified. The results from the step trend test
also mainly identified a positive relationship between the tree cover
and the rainfall at p \textless{} 0.1 at the NSW/Victoria region. High
rainfall variability and possible regrowth could have impacted the
results in the Queensland region.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

\begin{document}

@@

In Australia, significant tree cover change has mainly occurred in the
north east and south east of the continent, as well as in the southwest
of Western Australia. According to the National Dynamic Land Cover
Dataset (DLCD) (Lymburner et al. 2010), most of these areas have
experienced decreases in the Enhanced Vegetation Index (EVI) post 2000,
as derived from satellite data. As an index for vegetation greenness,
the decreasing values of EVI indicate lower biomass over time in the
tree cover regions. The possible EVI reduction might be due to land
clearing, bush fires or drought.

@@

Two regions were selected where significant tree cover change since 2000
was reported. The first region is located in south central Queensland
(QLD) partly covering the north of the Murray Darling Basin (MDB) (site
1 in Figure \ref{fig:selreg}). High rates of land clearing have been
reported in this region during the early 2000s (Department of Natural
Resources and Water 2007). The second study region is located at the
border of New South Wales and Victoria (NSW/VIC), and includes the Snowy
Mountain ranges (site 2 in Figure \ref{fig:selreg}). Severe bush fires
occurred in this area in early 2003 (see Figure \ref{fig:bushfire}). The
2003 bush fires were the largest and the worst in the last 60 years (The
State Government of Victoria 2011). Two thirds of Kosciuszko national
park was heavily burned and regrowth was reported to be slow due to
drought and cold conditions (ABC News 2003), and the type of species in
this region. However, in the longer term, after an early high
transpiration period a recovery of pre-fire evapotranspiration would be
expected (Kuczera 1987). For the purpose of this study, significant tree
cover loss has happened in both study areas in the last decade, either
permanently or temporarily.

@@

The land use and land cover characteristics in the two regions are also
different. In the Queensland region, the tree cover is sparse over most
of the area. The MODIS satellite tree cover data (discussed in more
detail in section 3) shows that tree cover in this region is generally
below 20\% of total ground area. Grazing is the main activity in this
region, with over 90\% of land used by the grazing industry (ABARES
2010). Our starting assumption is that the main cause of the EVI decline
over large part of the region is due to land clearing. Tree cover has
been cleared at a massive scale over the last decade, especially during
2002 - 2004. The reports from the Queensland Statewide Land Cover and
Trees Study (SLATS) (e.g. Department of Natural Resources and Mines
2005; Department of Science, Information Technology and Innovation 2017)
were used to investigate the time and location of the land clearing in
the QLD region.

The Kosciuszko national park is within the NSW/VIC region. Here tree
cover is denser with open or even closed forest (the tree cover
distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant
species in the alpine area are Snow Gum and large stand species such as
Alpine Ash and Mountain Gum in the sub-alpine area. These trees can
reach a great height but they take long time to grow. For example,
Alpine Ash would need about 20 years to mature. Although land clearing
is not a major issue in this region, it is vulnerable to fires and
drought. The MODIS burned area product, MCD45A1 (Roy, Lewis, and Justice
2002; Roy et al. 2005; Roy et al. 2008), was used to locate bush fires
areas in the NSW/VIC region, with a grid resolution of 500 m. MCD45A1
provides monthly burning information on all pixels, which helps to
pinpoint an abrupt event.

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/tc_simple} \caption{The expected evolution of the land surface after trees have been removed in (a) the QLD region and (b) the NSW/VIC region.}\label{fig:figure3-tc-simple}
\end{figure}

The overall hypothesis is that the effect of 2003 - 2004 land clearings
in the QLD region and the 2003 bush fires in the NSW/VIC region cause a
step change in the local rainfall. The actual tree cover change during
this time at the pixel level was derived from the 15-year MODIS data
(discussed below). The difference of tree cover before and after the
land disturbance was tested using a Student's t-test. As the length of
the tree cover data is shorter than available rainfall data, earlier
land clearing in the QLD region cannot be identified spatially, hence
they are excluded from the analysis.

@@

Several land surface data sets were used in this study. The main one was
the MOD44B product Global Vegetation Continuous Field data set (version
5). This data set provides estimates of percent tree cover (percentage
of ground surface covered by trees) at a grid resolution of 250 m
(Townshend et al. 2011), which is finer than the earlier mentioned
burned product MCD45A1. The data set is available on an annual time
interval for the study period of 2000 - 2015. The tree cover data was
produced from 16-day Terra MODIS Land Surface Reflectance data and Land
Surface Temperature (Townshend et al. 2011). The National Dynamic Land
Cover Dataset (DLCD) (Lymburner et al. 2010) from the Australian
Collaborative Land Use Mapping Program (ACLUMP) was used to verify the
trend of vegetation cover change calculated from the previous data set.
This data set, developed by Geoscience Australia and Australian Bureau
of Agricultural and Resource Economics and Sciences (ABARES), is the
first nationally consistent and thematically comprehensive land cover
reference for Australia. The DLCD is based on the 16-day Enhanced
Vegetation Index (EVI), again from the MODIS satellite, between April
2000 and April 2015. It also has a grid resolution of 250 m. The data
set provides information on the final land cover types (as in 2015) and
estimated trend of EVI statistics (annual mean, maximum and minimum).

The SILO Rainfall product data for Australia was used (Jeffrey et al.
2001) (available online at \url{https://silo.longpaddock.qld.gov.au/}))
. The data has been projected onto a national 0.05\textdegree  by
0.05\textdegree  grid (approximately 5 km by 5 km). This gridded data
set was generated from station observations using spline interpolation
and kriging (Jeffrey et al. 2001). The data has been compared to other
gridded products and observed data and is generally of high quality
(Tozer, Kiem, and Verdon-Kidd 2009; Tozer, Kiem, and Verdon-Kidd 2012).
The data is available on a daily and monthly basis from 1889 to current.
Here a subset of 36 years (1979 - 2015) was used. The study was
conducted on monthly data, as a land cover change effect on annual
rainfall might be negligible, but can often be significant in particular
months or seasons (e.g. Otterman et al. 1990; Gaertner et al. 2001;
Semazzi and Song 2001; Oleson et al. 2004; Deo et al. 2009).

Large scale climate drivers are represented by various climatic indices.
The Southern Oscillation Index (SOI) is generally regarded as a good
predictor of Australian rainfall (Risbey et al. 2009; Chowdhury and
Beecham 2010; Westra and Sharma 2010), but its skill is weaker in some
parts of Australia. For example the Southern Annular Mode (SAM) is found
to be more important than ENSO in south Western Australia (Meneghini,
Simmonds, and Smith 2007). The testing of the suitability of each index
for the regions of interest is described in a later section. The
following climate indices were used as candidate predictors for local
rainfall.

\begin{itemize}
\tightlist
\item
  Southern Oscillation Index (SOI). The Troup version of the monthly SOI
  series used in this study was obtained from BoM (available online at
  \url{http://www.bom.gov.au/climate/current/soihtm1.shtml}).\\
\item
  Eastern, East Central and Central Tropical Pacific Sea Surface
  Temperatures (NINO 3, NINO 3.4 and NINO 4). Monthly SST anomalies are
  available from IRI/LDEO data library and the extended NINO data set is
  used (available online at
  \url{http://iridl.ldeo.columbia.edu/SOURCES/.Indices/.nino/.EXTENDED/}).\\
\item
  Pacific Decadal Oscillation (PDO). The Pacific Decadal Oscillation is
  the leading principal component of monthly SST anomaly in the North
  Pacific Ocean.. The monthly PDO series was provided by JISAO (Joint
  Institute for the Study of the Atmosphere and Ocean, University of
  Washington) (available online at
  \url{http://jisao.washington.edu/pdo/PDO.latest}).\\
\item
  Indian Ocean Dipole (IOD). The Indian Ocean dipole is commonly
  measured by the difference between SST anomaly in the western (50 -
  70\textdegree E and 10\textdegree S-10\textdegree N) and eastern (90 -
  110\textdegree E and 0 - 10\textdegree S) equatorial India Ocean (Saji
  et al. 1999). Monthly IOD was obtained from JAMSTEC (the Japan Agency
  for Marine-Earth Science and Technology) (available online at
  \url{http://www.jamstec.go.jp/frcgc/research/d1/iod/DATA/dmi.monthly.txt}).
\end{itemize}

@@

As highlighted in the introduction, the Australian climate is influenced
by sea surface temperatures in the tropical Pacific and Indian Oceans,
as well as pressure systems in the Southern Ocean (BoM 2012b). Risbey et
al. (2009) compared five large-scale drivers, including ENSO (measured
by SOI and the Tropical Pacific Sea surface temperatures (SSTs)), IOD,
SAM, MJO (Madden-Julian oscillation) and blocking, in relation to
Australian rainfall variability. The MJO is a large scale
eastward-propagating wave-like disturbance located around equatorial
latitudes (Risbey et al. 2009). They identified SOI as the most
important index among all climate indices tested for broad parts of
Australia (including QLD and NSW/VIC) in almost any season. In this
study, four climate indices were selected from the main climatic
indicators (see section \protect\hyperlink{Data}{Data}) and used as the
explanatory variables in the model for each study region. A further
complicating factor is the influence of the ``millenium drought'' over
the study period and in particular the change to wet conditions in 2010
- 2011 (Dijk and Viney 2013). Therefore, the spatially averaged monthly
rainfall in the Murray Darling Basin (MDB, downloaded from
\href{http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt}{the
Bureau of Meteorology}) was used to explain the year-on-year variation
in the rainfall in the regions. As climate alone are unlikely able to
fully explain the temporal variation in rainfall in the regions,
including the drought (Dijk and Viney 2013; Westra and Sharma 2010),
this variable was included. Since both regions at least partly overlap
with the MDB, the average rainfall for the entire basin was assumed to
be a useful explaining variable.

Correlations between rainfall and each climate index were analysed.
Rainfall in each study region was first deseasonalised and detrended
using the seasonal decomposition function \texttt{ds} in the package
\texttt{deseasonalise} in R (R Core Team, 2018). Using detrended data
gives a better indication of the underlying correlation by removing the
underlying correlation in the data (Smith and Timbal 2012). The
cross-correlations between the deseasonalised and detrended rainfall and
the climatic indices were tested using the Pearson's product moment
correlation method, assuming the relationships are linear. Although the
optimal technique for exploring the correlation with each index could be
different as described in Risbey et al. (2009), the Pearson's method was
applied to all indices for consistency. Because the PDO describes the
multi-decadal SST with lower frequency (MacDonald and Case 2005;
Zanchettin et al. 2008; Kamruzzaman, Beecham, and Metcalfe 2011),
instead of 37-year rainfall data, a longer period (115 years, from 1900
to 2015) was used to estimate the correlation with PDO, up to lag 24.
For the other indices, the 37-year data was used.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/cor_qld} \caption{Cross-correlation of six climate indices and rainfall in QLD study region. For the PDO analysis, 108-year rainfall data (1900 - 2008) are used. Otherwise, 36-year rainfall data are used. The correlation with NINO 3 is not shown as it is very similar to but weaker than for NINO 3.4.}\label{fig:cor-rain-qld}
\end{figure}

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/cor_nswvic} \caption{Cross-correlation of six climate indices and rainfall in NSW/VIC study region.}\label{fig:cor-rain-nsw}
\end{figure}

Based on the correlation between the climatic indices and rainfall in
the regions (as shown in Figure \ref{fig:cor-rain-qld} and Figure
\ref{fig:cor-rain-nsw}), it can be concluded that:

\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.
  \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
  \item In both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
  \item In general, for the better correlated indices, strongest correlations occured at zero time lags.
\end{itemize}

The above findings are consistent with previous studies. Although some
indices are serially correlated with rainfall up to several months, the
lag zero events have the greatest correlation coefficients. Using
multiple climatic index series was generally found most useful in
rainfall prediction (e.g. Risbey et al. 2009; Kamruzzaman, Beecham, and
Metcalfe 2011).

Rainfall in Australia shows strong seasonal patterns (Holper 2011;
Australian Bureau of Statistics 2012). For example, the north part of
the country is summer rainfall dominant with a dry winter, while most of
the southern part has a winter rainfall regime. This character is driven
by the movement of subtropical high pressure systems which dominate the
Australian climate (BoM 2012a). These characteristic summer and winter
rainfall patterns mean that the seasonal component of rainfall has a
periodic pattern which should be included in the model.

Long term trends in the regional rainfall in some parts of Australia are
significant (Hughes 2003; Gallant, Hennessy, and Risbey 2007; Chowdhury
and Beecham 2010). In the northern and eastern parts of the continent,
increasing rainfall is reported over the last century (Hughes 2003). The
presence of such long term trends may be confused with the outcome of a
step change in rainfall. As a result a linear trend term was implemented
in the model to remove any long term effect.

We assumed all the factors are additive smooth components in determining
rainfall following Kamruzzaman, Beecham, and Metcalfe (2011). In this
case, the rainfall model is a generalised additive model (GAM) (Hastie
and Tibshirani 1986) with a log link function \texttt{g()} and assuming
the residuals are gamma distributed (see Figure \ref{fig:hist-rain}).
This means all predictors are modelled as smooth functions, in this case
using the shrinkage version of the cubic regression splines (Wood 2011).
All splines were limited to 3 knots in flexibility (Wood 2011), to
reduce the risk of overfitting. \vspace{0.5cm}

\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}} + s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\
&s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{Season}) + \\
           &\beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
\end{array}
\label{eq:model}
\end{equation}

The bold letters represent the time series vectors. The region is
indicated by \(r\), while \(\beta_u\) (\(_u\)=0, 1) are the fitted
coefficients in the model. \(s_v\) (\(_v\)=1, 2, 3,\ldots{}) are the
smooth penalized cubic regression spline functions on the climatic
indices and the season. MDB\textsubscript{monthlyRain} is the spatially
averaged monthly rainfall across the Murray Darling Basin. Apart from
dropping PDO as a predictor, all other climate indices were included,
allowing the model to select the appropriate predictors.

A possible linear long term trend in the rainfall data is modelled by
\textbf{Trend} = 1,2,3\ldots{}n, where n is the total number of months
in the time series. \textbf{Season} is the seasonal component. The
climatic terms are also modelled with smooth functions. The effect of
large scale drivers on Australian rainfall is more likely to be seasonal
(Murphy and Timbal 2008; Schepen, Wang, and Robertson 2012), and the
smooth spline function is more flexible in reproducing the variability
in impacts of the climatic indices.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/hist_rainfall} \caption{Distribution of monthly rainfall in (a) QLD and (b) NSW/VIC. Using a Kolmogorov-Smirnov test with shape = 1 and 2.4  for QLD and NSW/VIC respectively, rainfall in both regions can be modelled as a gamma distribution.}\label{fig:hist-rain}
\end{figure}

@@

\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{MDB_{monthlyRain}} + \\
  &s'_2(\mathbf{SOI}) + s'_3(\mathbf{IOD}) + \\
  &s'_4(\mathbf{Nino3.4}) + s'_5(\mathbf{Nino4}) + s'_6\mathbf{Season} + \\
  &\beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
\end{array}
\label{eq:model2}
\end{equation}

@@

To support the regression analysis, a Mann-Whitney Rank-Sum step trend
test was used to detect changes in rainfall as a result of vegetation
cover change. This specific nonparametric statistical test was modified
from the Mann-Whitney U test by Hirsch and Gilroy (1985) and can
identify a step change in data which is cross-correlated. The gridded
rainfall dataset used in this study has a high spatial correlation
between neighbouring pixels due to the underlying interpolation method.
The advantages of using the Rank-Sum test are: (1) it does not depend on
assumptions of the data distribution; (2) it is not restricted to
datasets with no missing data; (3) it is robust and not as easily
influenced by outliers and negative numbers (Hirsch and Gilroy 1985).
However, the test has less power than parametric tests.

The rainfall residuals from the regression model in Equation
\eqref{eq:model2} were used for the step trend test. To detect a step
change, using deseasonalised and detrended data is important (Hirsch and
Gilroy 1985). Furthermore, as rainfall can only be partially attributed
to local sources and conditions, other effects are introduced by large
scale dynamics and changes in climatic factors. The assumption is that
the regression model should remove these effects, and additionally
deseasonalise and detrend the rainfall data. As a result, the local
landuse effects are amplified in the remaining residuals. The test,
described in the following sub section, subsequently associates trends
in the rainfall residuals with tree cover changes.

\subsubsection{Mann-Whitney rank-sum
statistic}\label{mann-whitney-rank-sum-statistic}

As indicated, the step trend test is a modified version of the
Mann-Whitney U statistic (Hirsch and Gilroy 1985). As a nonparametric
rank-based test, the Mann-Whitney test does not use the exact values of
rainfall but depends on the ranks of the data. For each month, rainfall
residuals of each year were ranked in an ascending order. The ranking of
January rainfall in a sample pixel k in QLD is illustrated below:

>>>>

\usepackage{lipsum} \usepackage{textcomp, rotating}
\usepackage[normalem]{ulem}

@@

\begin{abstract}
Analysis of observational data to pinpoint impact of land cover change
on local rainfall is difficult due to multiple environmental factors
that cannot be strictly controlled. In this study we use a statistical
approach to identify the relationship between removal of tree cover and
rainfall with data from best available sources for two large areas in
Australia. Gridded rainfall data between 1979 and 2015 was used for the
areas, while large scale (exogenous) effects were represented by mean
rainfall across a much larger area and climatic indicators, such as
Southern Oscillation Index and Indian Ocean Dipole. Both generalised
additive modelling and step trend tests were used for the analysis. For
a region in south central Queensland, the reported change in tree
clearing between 2002 - 2005 did not result in strong statistically
significant precipitation changes. On the other hand, results from a
bushfire affected region on the border of New South Wales and Victoria
suggests significant changes in the rainfall due to changes in tree
cover. This indicates the method works better when an abrupt change in
the data can be clearly identified. The results from the step trend test
also mainly identified a positive relationship between the tree cover
and the rainfall at p \textless{} 0.1 at the NSW/Victoria region. High
rainfall variability and possible regrowth could have impacted the
results in the Queensland region.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

@@

In Australia, significant tree cover change has mainly occurred in the
north east and south east of the continent, as well as in the southwest
of Western Australia. According to the National Dynamic Land Cover
Dataset (DLCD) (Lymburner et al., 2010), most of these areas have
experienced decreases in the Enhanced Vegetation Index (EVI) post 2000,
as derived from satellite data. As an index for vegetation greenness,
the decreasing values of EVI indicate lower biomass over time in the
tree cover regions. The possible EVI reduction might be due to land
clearing, bush fires or drought.

@@

Two regions were selected where significant tree cover change since 2000
was reported. The first region is located in south central Queensland
(QLD) partly covering the north of the Murray Darling Basin (MDB) (site
1 in Figure \ref{fig:selreg}). High rates of land clearing have been
reported in this region during the early 2000s (Department of Natural
Resources and Water, 2007). The second study region is located at the
border of New South Wales and Victoria (NSW/VIC), and includes the Snowy
Mountain ranges (site 2 in Figure \ref{fig:selreg}). Severe bush fires
occurred in this area in early 2003 (see Figure \ref{fig:bushfire}). The
2003 bush fires were the largest in the last 60 years (The State
Government of Victoria, 2011). Two thirds of Kosciuszko national park
was heavily burned and regrowth was suggested to be slow due to drought
and cold conditions (ABC News, 2003), and the type of species in this
region. However, in the longer term, after an early high transpiration
period a recovery of pre-fire evapotranspiration would be expected
(Kuczera, 1987). For the purpose of this study, significant tree cover
loss has happened in both study areas in the last decade, either
permanently or temporarily.

@@

The land use and land cover characteristics in the two regions are also
different. In the Queensland region, the tree cover is sparse over most
of the area. The MODIS satellite tree cover data (discussed in more
detail in section 3) shows that tree cover in this region is generally
below 20\% of total ground area. Grazing is the main activity in this
region, with over 90\% of land used by the grazing industry (ABARES,
2010). Our starting assumption is that the main cause of the EVI decline
over large part of the region is due to land clearing. Tree cover has
been cleared at a massive scale over the last decade, especially during
2002 - 2004. The reports from the Queensland Statewide Land Cover and
Trees Study (SLATS) (e.g. Department of Natural Resources and Mines,
2005; Department of Science, Information Technology and Innovation,
2017) were used to investigate the time and location of the land
clearing in the QLD region.

The Kosciuszko national park is within the NSW/VIC region. Here tree
cover is denser with open or even closed forest (the tree cover
distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant
species in the alpine area are Snow Gum (\emph{Eucalyptus
Pauciflora})and large stand species such as Alpine Ash (\emph{Eucalyptus
delegatensis}) and Mountain Gum (\emph{Eucalyptus dalrympleana}) in the
sub-alpine area. These trees can reach a great height but they take long
time to grow. For example, Alpine Ash (\emph{Eucalyptus delegatensis})
would need about 20 years to mature (40 - 45 m, Buckley et al. (2012)).
This region is vulnerable to fires and drought, however land clearing is
not a major issue. The MODIS burned area product, MCD45A1 (Roy, Lewis \&
Justice, 2002; Roy et al., 2005, 2008), was used to locate bush fires
areas in the NSW/VIC region, with a grid resolution of 500 m. MCD45A1
provides monthly burning information on all pixels, which helps to
pinpoint an abrupt event.

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/tc_simple} \caption{The expected evolution of the land surface after trees have been removed in (left) the QLD region and (right) the NSW/VIC region.}\label{fig:figure3-tc-simple}
\end{figure}

The overall hypothesis is that the effect of 2003 - 2004 land clearings
in the QLD region and the 2003 bush fires in the NSW/VIC region cause a
step change in the local rainfall. The actual tree cover change during
this time at the pixel level was derived from the 15-year MODIS data
(discussed below). As the length of the tree cover data is shorter than
available rainfall data, earlier land clearing in the QLD region cannot
be identified spatially, hence they are excluded from the analysis.

@@

Several land surface data sets were used in this study. The main one was
the MOD44B product Global Vegetation Continuous Field data set (version
5). This data set provides estimates of percent tree cover (percentage
of ground surface covered by trees) at a grid resolution of 250 m
(Townshend et al., 2011), which is finer than the earlier mentioned
burned product MCD45A1. The data set is available on an annual time
interval for the study period of 2000 - 2015. The tree cover data was
produced from 16-day Terra MODIS Land Surface Reflectance data and Land
Surface Temperature (Townshend et al., 2011). The National Dynamic Land
Cover Dataset (DLCD) (Lymburner et al., 2010) from the Australian
Collaborative Land Use Mapping Program (ACLUMP) was used to verify the
trend of vegetation cover change calculated from the previous data set.
This data set, developed by Geoscience Australia and Australian Bureau
of Agricultural and Resource Economics and Sciences (ABARES), is the
first nationally consistent and thematically comprehensive land cover
reference for Australia. The DLCD is based on the 16-day Enhanced
Vegetation Index (EVI), again from the MODIS satellite, between April
2000 and April 2015. It also has a grid resolution of 250 m. The data
set provides information on the final land cover types (as in 2015) and
estimated trend of EVI statistics (annual mean, maximum and minimum).

The SILO Rainfall product data for Australia was used (Jeffrey et al.,
2001) (available online at \url{https://silo.longpaddock.qld.gov.au/}))
. The data has been projected onto a national 0.05\textdegree  by
0.05\textdegree  grid (approximately 5 km by 5 km). This gridded data
set was generated from station observations using spline interpolation
and kriging (Jeffrey et al., 2001). The data has been compared to other
gridded products and observed data and is generally of high quality
(Tozer, Kiem \& Verdon-Kidd, 2009, 2012). The data is available on a
daily and monthly basis from 1889 to current. Here a subset of 36 years
(1979 - 2015) was used. The study was conducted on monthly data, as a
land cover change effect on annual rainfall might be negligible, but can
often be significant in particular months or seasons (e.g. Otterman et
al., 1990; Gaertner et al., 2001; Semazzi \& Song, 2001; Oleson et al.,
2004; Deo et al., 2009).

Large scale climate drivers are represented by various climatic indices.
The Southern Oscillation Index (SOI) is generally regarded as a good
predictor of Australian rainfall (Risbey et al., 2009; Chowdhury \&
Beecham, 2010; Westra \& Sharma, 2010), but its skill is weaker in some
parts of Australia. For example the Southern Annular Mode (SAM) is found
to be more important than ENSO in south Western Australia (Meneghini,
Simmonds \& Smith, 2007). The testing of the suitability of each index
for the regions of interest is described in a later section. The
following climate indices were used as candidate predictors for local
rainfall.

\begin{itemize}
\tightlist
\item
  Southern Oscillation Index (SOI). The Troup version of the monthly SOI
  series used in this study was obtained from BoM (available online at
  \url{http://www.bom.gov.au/climate/current/soihtm1.shtml}).\\
\item
  Eastern, East Central and Central Tropical Pacific Sea Surface
  Temperatures (NINO 3, NINO 3.4 and NINO 4). Monthly SST anomalies are
  available from IRI/LDEO data library and the extended NINO data set is
  used (available online at
  \url{http://iridl.ldeo.columbia.edu/SOURCES/.Indices/.nino/.EXTENDED/}).\\
\item
  Pacific Decadal Oscillation (PDO). The Pacific Decadal Oscillation is
  the leading principal component of monthly SST anomaly in the North
  Pacific Ocean.. The monthly PDO series was provided by JISAO (Joint
  Institute for the Study of the Atmosphere and Ocean, University of
  Washington) (available online at
  \url{http://jisao.washington.edu/pdo/PDO.latest}).\\
\item
  Indian Ocean Dipole (IOD). The Indian Ocean dipole is commonly
  measured by the difference between SST anomaly in the western (50 -
  70\textdegree E and 10\textdegree S-10\textdegree N) and eastern (90 -
  110\textdegree E and 0 - 10\textdegree S) equatorial India Ocean (Saji
  et al., 1999). Monthly IOD was obtained from JAMSTEC (the Japan Agency
  for Marine-Earth Science and Technology) (available online at
  \url{http://www.jamstec.go.jp/frcgc/research/d1/iod/DATA/dmi.monthly.txt}).
\end{itemize}

@@

As highlighted in the introduction, the Australian climate is influenced
by sea surface temperatures in the tropical Pacific and Indian Oceans,
as well as pressure systems in the Southern Ocean (BoM, 2012). Risbey et
al. (2009) compared five large-scale drivers, including ENSO (measured
by SOI and the Tropical Pacific Sea surface temperatures (SSTs)), IOD,
SAM, MJO (Madden-Julian oscillation) and blocking, in relation to
Australian rainfall variability. The MJO is a large scale
eastward-propagating wave-like disturbance located around equatorial
latitudes (Risbey et al., 2009). They identified SOI as the most
important index among all climate indices tested for broad parts of
Australia (including QLD and NSW/VIC) in almost any season. In this
study, four climate indices were selected from the main climatic
indicators (see section \protect\hyperlink{Data}{Data}) and used as the
explanatory variables in the model for each study region. A further
complicating factor is the influence of the ``millenium drought'' over
the study period and in particular the change to wet conditions in 2010
- 2011 (Dijk \& Viney, 2013). Therefore, the spatially averaged monthly
rainfall in the Murray Darling Basin (MDB, downloaded from
\href{http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt}{the
Bureau of Meteorology}) was used to explain the year-on-year variation
in the rainfall in the regions. Since both regions at least partly
overlap with the MDB, the average rainfall for the entire basin was
assumed to be a useful explaining variable.

Correlations between rainfall and each climate index were analysed.
Rainfall in each study region was first deseasonalised and detrended
using the seasonal decomposition function \texttt{ds} in the package
\texttt{deseasonalise} in R (R Core Team, 2018). Using detrended data
gives a better indication of the underlying correlation by removing the
underlying correlation in the data (Smith \& Timbal, 2012). The
cross-correlations between the deseasonalised and detrended rainfall and
the climatic indices were tested, and the strongests indicators at lag 0
were identified for the model. Because the PDO describes the
multi-decadal SST with lower frequency (MacDonald \& Case, 2005;
Zanchettin et al., 2008; Kamruzzaman, Beecham \& Metcalfe, 2011),
instead of 37-year rainfall data, a longer period (115 years, from 1900
to 2015) was used to estimate the correlation with PDO, up to lag 24.
For the other indices, the 37-year data was used.

Rainfall in Australia shows strong seasonal patterns (Holper, 2011;
Australian Bureau of Statistics, 2012). As a result a seasonal component
of rainfall has a periodic pattern which should be included in the
model. In addition, long term trends in the regional rainfall in some
parts of Australia are significant (Hughes, 2003; Gallant, Hennessy \&
Risbey, 2007; Chowdhury \& Beecham, 2010). The presence of long term
trends can be confused with the outcome of a step change in rainfall. As
a result a linear trend term was implemented in the model to remove any
long term effects.

We assumed all the factors are additive smooth components in determining
rainfall following Kamruzzaman, Beecham \& Metcalfe (2011). In this
case, the rainfall model is a generalised additive model (GAM) (Hastie
\& Tibshirani, 1986) with a log link function \texttt{g()} and assuming
the residuals are gamma distributed. We used the shrinkage version of
the cubic regression splines (Wood, 2011) as smooth function. All
splines were limited to 3 knots in flexibility (Wood, 2011), to reduce
the risk of overfitting. \vspace{0.5cm}

\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta_0 + s_1(\mathbf{MDB_{monthlyRain}}) + s_2(\mathbf{SOI}) + s_3(\mathbf{IOD}) + \\
&s_4(\mathbf{Nino3.4}) + s_5(\mathbf{Nino4}) + s_6(\mathbf{PDO}) + \\
&s_7(\mathbf{Season}) + \beta_1\mathbf{Trend} + \boldsymbol{\epsilon}_r
\end{array}
\label{eq:model}
\end{equation}

The bold letters represent the time series vectors. The region is
indicated by \(r\), while \(\beta_u\) (\(_u\)=0, 1) are the fitted
coefficients in the model. \(s_v\) (\(_v\)=1, 2, 3,\ldots{}) are the
smooth penalized cubic regression spline functions on the climatic
indices and the season. MDB\textsubscript{monthlyRain} is the spatially
averaged monthly rainfall across the Murray Darling Basin.

The variable \textbf{Trend} = 1,2,3\ldots{}n, where n is the total
number of months in the time series is the possible long term trend in
the data, while \textbf{Season} is the seasonal component. The climatic
terms are also modelled with smooth functions as the effect of large
scale drivers on Australian rainfall can be seasonal (Murphy \& Timbal,
2008; Schepen, Wang \& Robertson, 2012).

@@

\begin{equation}
\begin{array}{lll}
g(E(\mathbf{R}_r)) = &\beta'_0 + s'_1(\mathbf{MDB_{monthlyRain}}) + \\
  &s'_2(\mathbf{SOI}) + s'_3(\mathbf{IOD}) + \\
  &s'_4(\mathbf{Nino3.4}) + s'_5(\mathbf{Nino4}) + s'_6\mathbf{PDO} + s'_7\mathbf{Season} + \\
  &\beta'_1\mathbf{Trend} + \beta'_2\mathbf{LC} + \boldsymbol{\epsilon'}_r
\end{array}
\label{eq:model2}
\end{equation}

@@

To support the regression analysis, a Mann-Whitney Rank-Sum step trend
test was used to detect changes in rainfall as a result of vegetation
cover change. This specific nonparametric statistical test was modified
from the Mann-Whitney U test by Hirsch \& Gilroy (1985) and can identify
a step change in data which is cross-correlated. In this case, this is
important as the gridded rainfall dataset has a high spatial correlation
between neighbouring pixels. The advantages of using the Rank-Sum test
are: (1) it does not depend on assumptions of the data distribution; (2)
it is not restricted to datasets with no missing data; (3) it is robust
and not as easily influenced by outliers and negative numbers (Hirsch \&
Gilroy, 1985). However, the test has less power than parametric tests.
As a nonparametric rank-based test, it depends on the ranks of the data.

The rainfall residuals from the regression model in Equation
\eqref{eq:model2} were used. The assumption is that the regression model
deseasonalises and detrends the rainfall data (Hirsch \& Gilroy, 1985),
amplifying the local landuse effects. For each month, rainfall residuals
of each year were ranked in an ascending order. The ranking of January
rainfall in a sample pixel k in QLD is illustrated in Table
\ref{tab:exampleRank}.
